---
title: "Breast Cancer TCGA Analysis"
output: 
  github_document:
    toc: true
---

# Dependencies

```{r results="hide"}
library(tidyverse)
library(ggforce)
library(recount)
library(pheatmap)
library(viridis)
library(useful)
library(RColorBrewer)
```

# Load the data
Gene-level expression data from the breast cancer TCGA cohort  was downloaded [here](http://duffel.rail.bio/recount/TCGA/rse_gene_breast.Rdata). I'll work with a local copy here.

```{r}
load("../data/rse_gene_breast.Rdata")
rse <- scale_counts(rse_gene, round=F)
exp <- assay(rse)
gene.data <- as.data.frame(rowData(rse))
```

# Grab expression values of gene set
We're using a manually selected gene list to explore in the data. Kind of an ugly way to grab a data frame of expression, but it works.

```{r}
sox10 <- filter(gene.data, symbol=="SOX10")$gene_id
her2 <- filter(gene.data, symbol=="ERBB2")$gene_id
esr1 <- filter(gene.data, symbol=="ESR1")$gene_id
pgr <- filter(gene.data, symbol=="PGR")$gene_id
lcn2 <- filter(gene.data, symbol=="LCN2")$gene_id
ceacam1 <- filter(gene.data, symbol=="CEACAM1")$gene_id
l1cam <- filter(gene.data, symbol=="L1CAM")$gene_id
slk <- filter(gene.data, symbol=="SLK")$gene_id
sox8 <- filter(gene.data, symbol=="SOX8")$gene_id
sox9 <- filter(gene.data, symbol=="SOX9")$gene_id
ar <- filter(gene.data, symbol=="AR")$gene_id
pip <- filter(gene.data, symbol=="PIP")$gene_id

dat <- data.frame(sox10=exp[sox10,],
                  her2=exp[her2,],
                  esr1=exp[esr1,],
                  pgr=exp[pgr,],
                  lcn2=exp[lcn2,],
                  ceacam1=exp[ceacam1,],
                  l1cam = exp[l1cam,],
                  slk = exp[slk,],
                  sox8 = exp[sox8,],
                  sox9 = exp[sox9,],
                  ar = exp[ar,],
                  pip = exp[pip,]
                  )
```

This data frame will be used for plotting gene-gene expression plots.

# Identify molecular subtypes of samples
Breast cancer is often classified based on the expression of the hormone receptors Her2 (ERBB2), Esr1, and Pr. We'll cluster patients based on their expression.

```{r, fig.width=2.25, fig.height=6}
dat.mat <- as.matrix(dat[,c(2,3,4)]) %>% log1p() %>% scale(scale=T, center=T)
dat.mat[dat.mat>=2] <- 2
dat.mat[dat.mat<=(-2)] <- -2
heatmap <- pheatmap(dat.mat, color=viridis(100),  scale="none", cluster_cols=F,
         cluster_rows=T, cutree_rows=3, show_rownames=F, clustering_method="ward.D2",
         file="../figs/receptor.heatmap.png",
         width=2.25, height=6)
plot(heatmap$gtable)
```

The patients seem to stratify nicely into triple-negative (TNBC), HER2+, and luminal clusters.

## Get cluster IDs for each sample
We can retrieve the cluster ID from the heatmap object

```{r}
clusters <- as.data.frame(cutree(heatmap$tree_row, k=3))
colnames(clusters) <- "Cluster"
clusters$sampleID <- rownames(clusters)
table(clusters$Cluster)
```

Kind of annoying that it isn't clear which cluster is which, but based on the size, it's clear that Cluster 1 = Luminal, 2 = TNBC, 3 = HER2+

Let's put this information into our data frame of expression values
```{r}
dat$Cluster <- factor(clusters$Cluster, levels=c(2,3,1))
```

## Plot gene by cluster

```{r}
colors <- brewer.pal(8, "Dark2")[c(3,2,1)]
```


Function to make life easier
```{r}
plot_gene <- function(gene) {
  plot <- ggplot(dat, aes(x=Cluster, y=log2(dat[,gene] + 1))) +
    geom_sina(size=0.75, alpha=0.5, aes(colour=Cluster)) +
    stat_summary(fun.y=median, fun.ymin=median, fun.ymax=median,
                 geom='crossbar', width=0.5) +
    xlab('Cluster') + ylab('log2(Counts+1)') +
    scale_x_discrete(labels=c('TNBC', 'HER2+', 'Luminal A/B')) +
    scale_colour_manual(values=colors) +
    theme_classic() +
    theme(axis.text=element_text(size=10, color='black'),
          axis.title=element_text(size=12),
          legend.position='none')
  ggsave(plot, file=paste0('../figs/', gene, '.cluster.pdf'),
         width=3.25, height=2.9)
  plot
}
```


And actually running it
```{r, fig.width=3.25, fig.height=2.9}
plot_gene('sox10')
plot_gene('her2')
plot_gene('esr1')
plot_gene('pgr')
plot_gene('lcn2')
plot_gene('ceacam1')
plot_gene('l1cam')
plot_gene('slk')
plot_gene('sox8')
plot_gene('sox9')
plot_gene('ar')
plot_gene('pip')
```


# Constructing a TNBC gene signature

We want to use an unbiased approach to see what genes are good markers of each of the clusters we identified. We'll just use an anova to look for genes with differences across the groups.

## Construct models
```{r}
run_anova <- function(x){
  model <- oneway.test(x~factor(clusters$Cluster))
  res <- data.frame(p.val = model$p.value,
                    f.stat = as.numeric(model$statistic))
  return(res)
}
```

Run the function down the matrix, collecting BH-corrected p-values for each gene. First we'll just log-transform the matrix so it's more normally distributed.

```{r}
exp <- exp[rowSums(exp)!=0,] #remove zero variance rows
exp <- log2(exp+1)
```

Run the test

```{r}
cluster.models <- apply(exp, 1, run_anova)
cluster.models <- do.call(rbind.data.frame, cluster.models)
cluster.models$gene_id <- rownames(cluster.models)
cluster.models <- left_join(cluster.models, gene.data, by="gene_id")
cluster.models$bp_length <- NULL
cluster.models$q.val <- p.adjust(cluster.models$p.val, method="BH")
cluster.models <- filter(cluster.models, symbol != "NA")
nrow(filter(cluster.models, q.val <= 0.01))
```

The joys of large sample sizes. We need an effect-size cutoff. 

## Calculating effect size
I'll calculate the average expression across clusters, and filter for genes with a high fold change between the highest and lowest values

```{r}
exp <- exp[cluster.models$gene_id,] #to remove the few rows that have NA as gene
```

```{r}
TNBC.samples <- filter(clusters, Cluster==2)$sampleID
HER2.samples <- filter(clusters, Cluster==3)$sampleID
Luminal.samples <- filter(clusters, Cluster==1)$sampleID
```

```{r}
exp.averages <- data.frame(TNBC = rowMeans(exp[,TNBC.samples]),
                           HER2 = rowMeans(exp[,HER2.samples]),
                           Luminal = rowMeans(exp[,Luminal.samples]))
exp.averages <- as.matrix(exp.averages)
cluster.models$diff <- rowMaxs(exp.averages) - rowMins(exp.averages)
```

```{r, fig.width=4, fig.height=3.5}
hist(cluster.models$diff, breaks=40, col="firebrick", pch=20,
     xlab="Exp. Difference", main="")
```

## Filtering by effect size
Cut off is really arbitrary here. We want to choose a cutoff that brings the marker list down to something manageable. We'll try markers

```{r}
sig.markers <- filter(cluster.models, q.val <= 0.01, diff >= 4)
nrow(sig.markers)
```

## Plotting marker genes

### Averages for each cluster

```{r, fig.width=2, fig.height=3.75}
marker.heatmap <- pheatmap(exp.averages[sig.markers$gene_id,],
                    color=magma(100),
                    scale="row",
                    cluster_rows=T,
                    cluster_cols=F,
                    cutree_rows=4,
                    show_rownames=F,
                    clustering_method="ward.D2",
                    filename="../figs/marker.heatmap.png",
                    width=2, height=3.75)
plot(marker.heatmap$gtable)
```

And just grab the cluster ID for each marker
```{r}
gene.clusters <- as.data.frame(cutree(marker.heatmap$tree_row, k=4))
colnames(gene.clusters) <- "Cluster"
gene.clusters$gene_id <- rownames(gene.clusters)
gene.clusters <- left_join(gene.clusters, gene.data, by="gene_id")
gene.clusters$Cluster <- factor(gene.clusters$Cluster)
sig.markers$Cluster <- gene.clusters$Cluster
summary(gene.clusters$Cluster)
```

Heatmap top to bottom: 2, 3, 4, 1

#### Where are specific markers?
We want to identify specifically where SOX10, AR, PGR, and HER2 are

```{r}
heatmap.avg.order <- marker.heatmap$tree_row$order
length(heatmap.avg.order)
```

##### SOX10
```{r}
which(sig.markers$gene_id[heatmap.avg.order] == sox10)
```

##### AR
```{r}
which(sig.markers$gene_id[heatmap.avg.order] == ar)
```

##### PGR
```{r}
which(sig.markers$gene_id[heatmap.avg.order] == pgr)
```

##### HER2
```{r}
which(sig.markers$gene_id[heatmap.avg.order] == her2)
```

### Across all samples

Getting the matrix ready
```{r}
exp.markers <- exp[sig.markers$gene_id,]
exp.markers <- t(scale(t(exp.markers), scale=T, center=T))
exp.markers[exp.markers > 1.25] <- 1.25
exp.markers[exp.markers < (-1.25)] <- -1.25
```

I want to use the ordering of the original heatmap used to subtype each sample. For the columns (genes), I'll order the columns simply by cluster ID, as ordered in the heatmap (ie. clusters 2, 3, 4, 1.

```{r}
sig.markers <- arrange(sig.markers, desc(diff)) #intragroup order by diff
gene.order <- c(filter(sig.markers, Cluster == 2)$gene_id,
                filter(sig.markers, Cluster == 3)$gene_id,
                filter(sig.markers, Cluster == 4)$gene_id,
                filter(sig.markers, Cluster == 1)$gene_id)
sample.order <- heatmap$tree_row$order
```

Set up annotation_row and annotation_col for pheatmap to identify what clusters things belong to
```{r}
annotation_row <- clusters
annotation_row$sampleID <- NULL
annotation_row$Cluster[annotation_row$Cluster==2] = "TNBC"
annotation_row$Cluster[annotation_row$Cluster==3] = "HER2+"
annotation_row$Cluster[annotation_row$Cluster==1] = "Luminal A/B"

annotation_col <- data.frame(MarkerCluster = factor(gene.clusters$Cluster))
rownames(annotation_col) <- gene.clusters$gene_id
```


```{r, fig.width=4, fig.height=5}
marker.exp.heatmap <- pheatmap(t(exp.markers[gene.order, sample.order]),
                               color=viridis(100),
                               cluster_rows=F,
                               cluster_cols=F,
                               scale="none",
                               show_rownames=F,
                               show_colnames=F,
                               annotation_col=annotation_col,
                               annotation_row=annotation_row,
                               filename="../figs/marker.heatmap.allsamples.png",
                               width=4, height=5)
plot(marker.exp.heatmap$gtable)
```

#### Where are specific markers in this heatmap

```{r}
which(gene.order==sox10)
which(gene.order==ar)
which(gene.order==pgr)
which(gene.order==her2)
```


#Gene-gene plots

The manuscript dives into androgen receptor (AR) expression patterns across molecular subtypes. I just want to make some plots comparing AR to other markers across each cluster. All this info is contained in the 'dat' data frame right now

## AR vs. Markers of clusters

```{r, fig.width=3.75, fig.height=2.9}
ar.sox10 <-  ggplot(dat, aes(x=log2(dat[,"ar"]+1), y=log2(dat[,"sox10"]+1))) +
  geom_point(size=1.5, alpha=0.5, aes(color=Cluster)) +
  geom_vline(xintercept=9, linetype="dashed") +
  geom_hline(yintercept=mean(log2(dat[,"sox10"] + 1)),
             linetype="dashed") +
  xlab("AR") + ylab("SOX10") +
  scale_colour_manual(values=colors[c(1,2,3)]) +
  scale_x_continuous(limits=c(3.65,15.5)) +
  scale_y_continuous(limits=c(0, max(log2(dat[,"sox10"]+1))+0.5)) +
  theme_classic() + 
  theme(axis.text=element_text(size=10, color='black'),
        axis.title=element_text(size=12))
ggsave(ar.sox10, file="../figs/ar.sox10.pdf", width=3.75, height=2.9)
ar.sox10

ar.her2 <-  ggplot(dat, aes(x=log2(dat[,"ar"]+1), y=log2(dat[,"her2"]+1))) +
  geom_point(size=1.5, alpha=0.5, aes(color=Cluster)) +
  geom_vline(xintercept=9, linetype="dashed") +
  geom_hline(yintercept=mean(log2(dat[,"her2"] + 1)),
             linetype="dashed") +
  xlab("AR") + ylab("HER2") +
  scale_colour_manual(values=colors[c(1,2,3)]) +
  scale_x_continuous(limits=c(3.65,15.5)) +
  scale_y_continuous(limits=c(0, max(log2(dat[,"her2"]+1))+0.5)) +
  theme_classic() + 
  theme(axis.text=element_text(size=10, color='black'),
        axis.title=element_text(size=12))
ggsave(ar.her2, file="../figs/ar.her2.pdf", width=3.75, height=2.9)
ar.her2

ar.pgr <-  ggplot(dat, aes(x=log2(dat[,"ar"]+1), y=log2(dat[,"pgr"]+1))) +
  geom_point(size=1.5, alpha=0.5, aes(color=Cluster)) +
  geom_vline(xintercept=9, linetype="dashed") +
  geom_hline(yintercept=mean(log2(dat[,"pgr"] + 1)),
             linetype="dashed") +
  xlab("AR") + ylab("PR") +
  scale_colour_manual(values=colors[c(1,2,3)]) +
  scale_x_continuous(limits=c(3.65,15.5)) +
  scale_y_continuous(limits=c(0, max(log2(dat[,"pgr"]+1))+0.5)) +
  theme_classic() + 
  theme(axis.text=element_text(size=10, color='black'),
        axis.title=element_text(size=12))
ggsave(ar.pgr, file="../figs/ar.pgr.pdf", width=3.75, height=2.9)
ar.pgr
```


##AR vs. Markers of clusters w/ subset

```{r, fig.width=3.25, fig.height=2.9}
dat.tnbc <- filter(dat, Cluster==2)
ar.tnbc <- ggplot(dat.tnbc, 
                  aes(x=log2(dat.tnbc[,"ar"]+1), y=log2(dat.tnbc[,"sox10"]+1))) + 
  geom_point(size=1.5, alpha=0.75, colour=colors[1]) +
  geom_vline(xintercept=9, linetype="dashed") +
  geom_hline(yintercept=mean(log2(dat[,"sox10"] + 1)),
             linetype="dashed") +
  xlab("AR") + ylab("SOX10") +
  scale_x_continuous(limits=c(3.65,15.5)) +
  scale_y_continuous(limits=c(0, max(log2(dat[,"sox10"]+1))+0.5)) +
  theme_classic() + 
  theme(axis.text=element_text(size=10, color='black'),
        axis.title=element_text(size=12))
ggsave(ar.tnbc, file="../figs/ar.sox10.tnbc.pdf", width=3.25, height=2.9)

dat.her2 <- filter(dat, Cluster==3)
ar.her2 <- ggplot(dat.her2, 
                  aes(x=log2(dat.her2[,"ar"]+1), y=log2(dat.her2[,"her2"]+1))) +
  geom_point(size=1.5, alpha=0.75, colour=colors[2]) +
  geom_vline(xintercept=9, linetype="dashed") +
  geom_hline(yintercept=mean(log2(dat[,"her2"] + 1)),
             linetype="dashed") +
  xlab("AR") + ylab("HER2") +
  scale_x_continuous(limits=c(3.65,15.5)) +
  scale_y_continuous(limits=c(0, max(log2(dat[,"her2"]+1))+0.5)) +
  theme_classic() +
  theme(axis.text=element_text(size=10, color='black'),
        axis.title=element_text(size=12))
ggsave(ar.her2, file="../figs/ar.her2.her2.pdf", width=3.25, height=2.9)

dat.luminal <- filter(dat, Cluster==1)
ar.luminal <- ggplot(dat.luminal, 
                  aes(x=log2(dat.luminal[,"ar"]+1),
                      y=log2(dat.luminal[,"pgr"]+1))) +
  geom_point(size=1.5, alpha=0.75, colour=colors[3]) +
  geom_vline(xintercept=9, linetype="dashed") +
  geom_hline(yintercept=mean(log2(dat[,"pgr"] + 1)),
             linetype="dashed") +
  xlab("AR") + ylab("PR") +
  scale_x_continuous(limits=c(3.65,15.5)) +
  scale_y_continuous(limits=c(0, max(log2(dat[,"pgr"]+1))+0.5)) +
  theme_classic() +
  theme(axis.text=element_text(size=10, color='black'),
        axis.title=element_text(size=12))
ggsave(ar.luminal, file="../figs/ar.pr.luminal.pdf", width=3.25, height=2.9)

ar.tnbc
ar.her2
ar.luminal
```

## Percentage of cells in "quadrants"

The paper discusses how genes that are highly expressed when AR is low is indicative of TNBC. We define an arbitrary (based on density) gate of AR expression at log2(counts+1)=9. The cutoff of the other markers is defined as half of the max log2(counts+1) value.

### TNBC

```{r}
#Top left
##Pass filter
nrow(filter(dat.tnbc, log2(ar + 1) < 9 & 
              log2(sox10 + 1) > mean(log2(dat[,"sox10"] + 1))))

##Percentage
nrow(filter(dat.tnbc, log2(ar + 1) < 9 & 
              log2(sox10 + 1) > mean(log2(dat[,"sox10"] + 1)))) / nrow(dat.tnbc)
```

```{r}
#Top right
##Pass filter
nrow(filter(dat.tnbc, log2(ar + 1) > 9 & 
              log2(sox10 + 1) > mean(log2(dat[,"sox10"] + 1))))

##Percentage
nrow(filter(dat.tnbc, log2(ar + 1) > 9 & 
              log2(sox10 + 1) > mean(log2(dat[,"sox10"] + 1)))) / nrow(dat.tnbc)
```

```{r}
#Bottom left
##Pass filter
nrow(filter(dat.tnbc, log2(ar + 1) < 9 & 
              log2(sox10 + 1) < mean(log2(dat[,"sox10"] + 1))))

##Percentage
nrow(filter(dat.tnbc, log2(ar + 1) < 9 & 
              log2(sox10 + 1) < mean(log2(dat[,"sox10"] + 1)))) / nrow(dat.tnbc)
```

```{r}
#Bottom right
##Pass filter
nrow(filter(dat.tnbc, log2(ar + 1) > 9 & 
              log2(sox10 + 1) < mean(log2(dat[,"sox10"] + 1))))

##Percentage
nrow(filter(dat.tnbc, log2(ar + 1) > 9 & 
              log2(sox10 + 1) < mean(log2(dat[,"sox10"] + 1)))) / nrow(dat.tnbc)
```

### HER2

```{r}
#Top left
##Pass filter
nrow(filter(dat.her2, log2(ar + 1) < 9 & 
              log2(her2 + 1) > mean(log2(dat[,"her2"] + 1))))

##Percentage
nrow(filter(dat.tnbc, log2(ar + 1) < 9 & 
              log2(her2 + 1) > mean(log2(dat[,"her2"] + 1)))) / nrow(dat.her2)
```

```{r}
#Top right
##Pass filter
nrow(filter(dat.her2, log2(ar + 1) > 9 & 
              log2(her2 + 1) > mean(log2(dat[,"her2"] + 1))))

##Percentage
nrow(filter(dat.her2, log2(ar + 1) > 9 & 
              log2(her2 + 1) > mean(log2(dat[,"her2"] + 1)))) / nrow(dat.her2)
```

```{r}
#Bottom left
##Pass filter
nrow(filter(dat.her2, log2(ar + 1) < 9 & 
              log2(her2 + 1) < mean(log2(dat[,"her2"] + 1))))

##Percentage
nrow(filter(dat.her2, log2(ar + 1) < 9 & 
              log2(her2 + 1) < mean(log2(dat[,"her2"] + 1)))) / nrow(dat.her2)
```

```{r}
#Bottom right
##Pass filter
nrow(filter(dat.her2, log2(ar + 1) > 9 & 
              log2(her2 + 1) < mean(log2(dat[,"her2"] + 1))))

##Percentage
nrow(filter(dat.her2, log2(ar + 1) > 9 & 
              log2(her2 + 1) < mean(log2(dat[,"her2"] + 1)))) / nrow(dat.her2)
```

### Luminal

```{r}
#Top left
##Pass filter
nrow(filter(dat.luminal, log2(ar + 1) < 9 & 
              log2(pgr + 1) > mean(log2(dat[,"pgr"] + 1))))

##Percentage
nrow(filter(dat.luminal, log2(ar + 1) < 9 & 
              log2(pgr + 1) > mean(log2(dat[,"pgr"] + 1)))) / nrow(dat.luminal)
```

```{r}
#Top right
##Pass filter
nrow(filter(dat.luminal, log2(ar + 1) > 9 & 
              log2(pgr + 1) > mean(log2(dat[,"pgr"] + 1))))

##Percentage
nrow(filter(dat.luminal, log2(ar + 1) > 9 & 
              log2(pgr + 1) > mean(log2(dat[,"pgr"] + 1)))) / nrow(dat.luminal)
```

```{r}
#Bottom left
##Pass filter
nrow(filter(dat.luminal, log2(ar + 1) < 9 & 
              log2(pgr + 1) < mean(log2(dat[,"pgr"] + 1))))

##Percentage
nrow(filter(dat.her2, log2(ar + 1) < 9 & 
              log2(pgr + 1) < mean(log2(dat[,"pgr"] + 1)))) / nrow(dat.luminal)
```

```{r}
#Bottom right
##Pass filter
nrow(filter(dat.luminal, log2(ar + 1) > 9 & 
              log2(pgr + 1) < mean(log2(dat[,"pgr"] + 1))))

##Percentage
nrow(filter(dat.luminal, log2(ar + 1) > 9 & 
              log2(pgr + 1) < mean(log2(dat[,"pgr"] + 1)))) / nrow(dat.luminal)
```


